---
title: "Class19"
author: "Yuntian Zhu (PID: A17816597)"
format: gfm
toc: true
---

## Background

Pertussis is a bacterial lung infection also known as whooping cough. Let's begin by examining CDCD reported case numbers in the US.

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L, 2024L),
        cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063, 22538)
       )
head (cdc)
```

We can check whether we can plot cases per year for pertussis in the US

```{r}
library(scales)
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = comma) +
  labs(x = "Year",
       y = "Number of cases")
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

Add some major milestone timepoints to out plot.

```{r}
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = comma) +
  labs(x = "Year",
       y = "Number of cases") +

  # vertical lines
  geom_vline(xintercept = 1947, color = "darkgreen") +
  geom_vline(xintercept = 1996, color = "darkred") +
  geom_vline(xintercept = 2020, color = "darkblue") +

  # lower labels (25% up the plot)
  annotate("text", x = 1947, y = max(cdc$cases)*0.25,
           label = "wP", color = "darkgreen", angle = 90, vjust = -0.5) +
  annotate("text", x = 1996, y = max(cdc$cases)*0.25,
           label = "aP", color = "darkred", angle = 90, vjust = -0.5) +
  annotate("text", x = 2020, y = max(cdc$cases)*0.25,
           label = "COVID-19", color = "darkblue", angle = 90, vjust = -0.5)
```

What I noticed:

The full introduction of mandatory wP (whole-cell) pertussis immunization in the mid 1940w leads to a dramatic reduction in case numbers (from over 200,000 to 100s).

The switch to the aP (never acellular vaccine) leads to an increasing number of cases, due to the fact that the immunity produced by aP is generally weaker than wP. There may also be other factors contributing to the increasing number of cases. 

The 2020 lockdowns and social distancing measures resulted in a decreasing number of cases.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the acellular pertussis (aP) vaccine in 1996, pertussis cases began rising again, with multiple outbreaks in the following decades. One likely explanation is that the aP vaccine produces weaker and more rapidly waning immunity compared to the older whole-cell vaccine, allowing more transmission and susceptibility in adolescents and adults. Additional contributors may include bacteria evolution and improved diagnostic testing using PCR. Social factors, including decreased vaccine uptake due to the modern anti-vaccine movement, may also amplify the resurgence by weakening herd immunity.

## The CMI-PB project

The CMI-PB project website: https://www.cmi-pb.org. The mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

They make their data available via JSON format API endpoints -- basically the database tables in a key::value type format. To read this, we can use the `read_jason()` function from the **jsonlite** package. 

```{r}
library (jsonlite)

subject <- read_json(path = "https://www.cmi-pb.org/api/v5_1/subject"
                     , simplifyVector = TRUE)

head (subject)
```

Q. How many individuals are in this dataset?

```{r}
nrow(subject)
```


> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

Q. What is the breakdown of race?

```{r}
table(subject$race)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

This breakdown is not particular representative of the US population.

## Investigating the Age Differences

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(dplyr)
library(lubridate)

subject <- subject %>%
  mutate(age = time_length(today() - ymd(year_of_birth), "years"))

subject %>%
  group_by(infancy_vac) %>%  
  summarise(avg_age = mean(age, na.rm = TRUE))
```

The average age of wP individuals is 36.57 years old. The average age of aP individuals is 27.82 years old. 

```{r}
x <- t.test(age ~ infancy_vac, data = subject)
x$p.value
```

A t-test indicates that the p value is 2.37e-23, far below the common 0.05 cutoff. Therefore, the ages of aP and wP individuals are statistically significantly different. 

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
t.test(age ~ infancy_vac, data = subject)
x$p.value
```

The aP and wP groups differ very strongly in age. The histogram shows that the aP participants cluster tightly around their mid-20s to early 30s, whereas the wP group is substantially older and has a much wider age range extending into the 40s and 50s. A formal Welch two-sample t-test confirms this visual impression, yielding an extremely small p-value (p = 2.37e-23), indicating that the age distributions of the two groups are statistically different.

## Joining Multiple Tables

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen"
                      , simplifyVector = T)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer"
                      , simplifyVector = T)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

We need to link these tables with the `subject` table, so we can begin to analyze the data and know who a given Ab sample was collected from and when. 

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(subject, specimen)
dim(meta)
head(meta)
```

Now, let's also join the `ab_titer` table with the `meta` table.

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
ab_data <- inner_join(meta, ab_titer)
dim(ab_data)
head(ab_data)
```

Q. How many Ab measurements do we have in total?

```{r}
nrow(ab_data)
```

Q. How many different isotypes (types of Ab) are in the dataset?

```{r}
unique(ab_data$isotype)
```

Q. How many different antigens?

```{r}
unique(ab_data$antigen)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(ab_data$isotype)
```

> Q12. What are the different $dataset values in ab_data and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(ab_data$dataset)
```


The dataset column contains four values: 2020_dataset, 2021_dataset, 2022_dataset, and 2023_dataset. The number of rows differs greatly across years: 2020 has by far the largest number of records (31,520), while 2022 and 2021 are much smaller (~7–8k). Interestingly, the most “recent” dataset (2023) does not have the largest number of rows -- it has only 15,050 entries, about half of 2020. This suggests that more recent data collection has been reduced, is still incomplete or reflects changes in sampling effort or assay prioritization.

## Examining IgG Ab Titer levels

```{r}
ggplot(ab_data) +
  aes(MFI, antigen) +
  geom_boxplot()
```

```{r}
igg <- ab_data |>
  filter(isotype == "IgG")
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

We can "facet" our plot by wP vs aP

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

Several pertussis-related antigens show clear changes in IgG levels over time, particularly PT, PRN, FHA and FIM2/3. These antigens rise noticeably after the early visits (around visits 4–7), showing boosted antibody responses following vaccination, and then gradually decline, reflecting antibody waning. In contrast, antigens like TT, DT and the negative control OVA remain relatively constant across all visits, showing no meaningful time-dependent change. This pattern occurs because only the pertussis antigens (PT, PRN, FHA and FIM2/3) are present in the Tdap booster dose administered during the study, whereas TT/DT were not boosted and OVA is unrelated to vaccination. As a result, only pertussis-specific antigens exhibit dynamic IgG responses over the 12-visit timeline.

We can also add some colors to the plots.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

More advanced analysis digging into individual antigen responses over time

```{r}
  filter(igg, antigen == "PT", dataset == "2021_dataset") |>
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
  geom_vline(xintercept=14, linetype = "dashed")
```

This plot compares PT-specific IgG responses after boosting between individuals primed in infancy with either whole-cell (wP) or acellular (aP) pertussis vaccines. Antibody levels rise rapidly and peak at 14 days post-boost (vertical dashed line), then gradually decline. Across the entire time course, wP-primed individuals show higher peak responses and sustain higher antibody levels than aP-primed individuals, demonstrating the more robust and durable recall immunity associated with wP priming.