---
title: "Analyzing Cloud Data"
author: "Yuntian Zhu (PID: A17816597)"
format: pdf
toc: true
---

## Data Import

First, let's input the analysis results that are done on AWS.

```{r}
library(tximport)
library(rhdf5)

# Make sure you are in the correct directory
setwd("/Users/apple")

# Find the quant directories
folders <- dir(pattern="_quant$")

# Extract sample names
samples <- sub("_quant", "", folders)

# Build the full paths to abundance.h5
files <- file.path(folders, "abundance.h5")
names(files) <- samples

# Import
txi.kallisto <- tximport(files, type="kallisto", txOut=TRUE)
```

We can take a look at our data.

```{r}
head(txi.kallisto$counts)
```

Let's find how many transcripts we have for each sample, and how many transcripts are detected in at least one sample

```{r}
colSums(txi.kallisto$counts)
```
```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

Then, let's do some data cleaning.

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## PCA

After importing the data and cleaning them, we can start the principal componenet analysis. 

```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```

Finally, let's make the PCA plots about PC1, PC2 and PC3 using the ggplot and ggrepel packages. 

```{r}
library(ggplot2)
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC2, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

```{r}
# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC1, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

```{r}
# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

ggplot(y) +
  aes(PC2, PC3, col=Condition) +
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_bw()
```

