---
title: "Class17"
author: "Yuntian Zhu (PID: A17816597)"
format: pdf
toc: true
---

## Section 1. Proportion of G/G in a population

Here we start by reading the downloaded CSV file.

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```


```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```

Now, let's look at a different population, GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

```{r}
table(gbr$Genotype..forward.strand.)/nrow(gbr)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.

Lets now dig into this further.

## Section 2. Population scale analysis (Homework)

Now, let's start some population scale analysis 

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
table(expr$geno)
```

Therefore, we have 108 A/A individuals, 233 A/G individuals and 121 G/G individuals.

```{r}
tapply(expr$exp, expr$geno, median)
```

Therefore, the median expression level for A/A is 31.24847, for A/G is 25.06486 and for G/G is 20.07363.





> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expr) + aes(geno, exp, fill = geno) +
  geom_boxplot(notch = TRUE)
```

The boxplot shows a clear, graded difference in ORMDL3 expression across the three genotypes. Individuals with the A/A genotype have the highest median expression, A/G individuals show an intermediate median and G/G individuals have the lowest median expression. The notches in the boxplot do not overlap substantially between A/A and G/G, indicating that their medians are significantly different.

Therefore, the SNP appears to have a strong effect on ORMDL3 expression: carrying the G allele is associated with decreased ORMDL3 expression, while the A allele is associated with higher expression.

